Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-31 doc/ApTranscriptome_manuscript.Rmd
Aborted attempt at re-ogranization of Intro

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew D. Nguyen, Mahesh Vangala, Jim Vincent, Robert R. Dunn,
  Aaron Ellison, Nate Sanders, Nicholas J. Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@cowles1939; @kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@deutsch2008; @kingsolver2013; @hoffmann2013] and these differences are often genetic [@krebs1997a; @kellermann2012]. To understand how this diversity of thermal tolerance evolved, we need more information on the genetic basis of thermal tolerance. Studies of the genetic basis of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998]. However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance. Little is known about the diversity of these genes or their expression patterns, especially in non-model terrestrial species.

@@

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??].

@@

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species. For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A. carolinensis*, or colder temperatures in *A. picea*. Additionally, genes may change their sensitivity to thermal extremes. Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

@@

Colonies of the ant species *Aphaenogaster picea* were collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis* from Durham, North Carolina ( 36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

>>>>

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: \centering{John Stanton-Geddes\textsuperscript{1}, Andrew D. Nguyen\textsuperscript{1}, Mahesh Vangala\textsuperscript{2}, James Vincent\textsuperscript{2}, Robert R. Dunn\textsuperscript{3}, Nathan Sanders\textsuperscript{4}, Nicholas J. Gotelli\textsuperscript{1}, Sara Helms Cahan\textsuperscript{1} \\ \textsuperscript{1}: Department of Biology, University of Vermont \\ \textsuperscript{2}: Vermont Genetics Network, University of Vermont \\ \textsuperscript{3}: North Carolina State University \\ \textsuperscript{4}: University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are often genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011].

. To understand how this variation of thermal tolerance evolved, we need information on the diversity of genes, and their expression patterns, involved in thermal tolerance.

In this study, we present a novel method for gene expression analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient, and report the set of genes that are thermally-responsive; the *thermal reactionome*. We then use a comparative approach to compare the thermal reactionomes between two closely-related ant species to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species.

@@

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species. For example, thermally-responsive genes may shift the temperature at which they are up-regulated to higher temperatures in *A. carolinensis*, or colder temperatures in *A. picea*. Additionally, genes may change their sensitivity to thermal extremes. Further, through the process of genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

@@

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012]. ...previous results from chambers...???

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.
